The very dependent recursive structure of iterated parametricity in indexed form

### 2025-12-29 paper/paper.tex
paper: s/Set/HSet/

Parametricity, like realisability, reducibility, and logical relations, was most commonly handled in an \emph{indexed} form, that is, using a type-theoretic language, considering predicates of type $A \rightarrow \Prop$, or families of type $A \rightarrow \Set$, or relations of type $A \times A \rightarrow \Prop$, or relevant relations -- sometimes called \emph{correspondences} -- of type $A \times A \rightarrow \Set$, where $\Prop$ and $\Set$ are types representing, that is classifying -- in category theory terminology --, propositions or sets, respectively. Along the indexed-fibered correspondence, that is along the equivalence between the types $A \rightarrow \Prop$ and $\Sigma B:\Set. (B \stackrel{inj}{\rightarrow} A)$ (proof-irrelevant case) and $A \rightarrow \Set$ and $\Sigma B:\Set. (B \rightarrow A)$ (proof-relevant case), for $A:\Set$, everything told in the indexed form can morally also be told in \emph{fibered} form. Such fibered approach to parametricity was for instance considered in Ghani \cite{Ghani...}, who also suggested to iterate parametricity.

@@

The internalisation of parametricity in unary indexed form was studied by \cite{bernardy12} directly leading to a type theory of iterated parametricity. Such type theory could be interpreted in a ``unary'' form of cubical set~\citep{bernardy15}, thus implicitly in fibered form since, in a binary cubical set (resp. in a unary cubical set and simplicial set as well), a set of segments over two points, that is a relevant relation, is given by first and second projection rather than as a function to $\Set$ (resp. a family is given as a \emph{total space} projected over a set of points $A:\Set$ rather than a family of type $A \rightarrow \Set$. This connection between iterated binary parametricity in fibered form and cubical sets includes an identification of degeneracies and reflexivity proofs, which can also be found in \cite{johann17}. In particular, it became clear at this point that iterated binary parametricity (without reflexivity) produces semi-cubical sets while it was also observed\footnote{The authors learned it from Hugo Moeneclaey and Thorsten Altenkirch.} that iterated unary parametricity produces augmented semi-simplicial sets.

>>>>

Parametricity, like realisability, reducibility, and logical relations, was most commonly handled in an \emph{indexed} form, that is, using a type-theoretic language, considering predicates of type $A \rightarrow \Prop$, or families of type $A \rightarrow \U$, or relations of type $A \times A \rightarrow \Prop$, or relevant relations -- sometimes called \emph{correspondences} -- of type $A \times A \rightarrow \U$, where $\Prop$ and $\U$ are types representing, that is classifying -- in category theory terminology --, propositions or sets, respectively. Along the indexed-fibered correspondence, that is along the equivalence between the types $A \rightarrow \Prop$ and $\Sigma B:\U. (B \stackrel{inj}{\rightarrow} A)$ (proof-irrelevant case) and $A \rightarrow \U$ and $\Sigma B:\U. (B \rightarrow A)$ (proof-relevant case), for $A:\U$, everything told in the indexed form can morally also be told in \emph{fibered} form. Such fibered approach to parametricity was for instance considered in Ghani \cite{Ghani...}, who also suggested to iterate parametricity.

@@

The internalisation of parametricity in unary indexed form was studied by \cite{bernardy12} directly leading to a type theory of iterated parametricity. Such type theory could be interpreted in a ``unary'' form of cubical set~\citep{bernardy15}, thus implicitly in fibered form since, in a binary cubical set (resp. in a unary cubical set and simplicial set as well), a set of segments over two points, that is a relevant relation, is given by first and second projection rather than as a function to $\U$ (resp. a family is given as a \emph{total space} projected over a set of points $A:\U$ rather than a family of type $A \rightarrow \U$. This connection between iterated binary parametricity in fibered form and cubical sets includes an identification of degeneracies and reflexivity proofs, which can also be found in \cite{johann17}. In particular, it became clear at this point that iterated binary parametricity (without reflexivity) produces semi-cubical sets while it was also observed\footnote{The authors learned it from Hugo Moeneclaey and Thorsten Altenkirch.} that iterated unary parametricity produces augmented semi-simplicial sets.
